### 2022-01-12 doc/src/AdvPLIC.tex
For Adv. PLIC, domaincfg.IE has no effect on extempore MSIs
Reduce complexity of Adv. PLIC's msiaddrcfg registers

Within a given interrupt domain, each of the domain's harts has a
unique \emph{index number} in the range 0 to ${\mbox{2}^{14}-\mbox{1}}$
(=~16,383).
The index number a domain associates with a hart may or may not have
any relationship to the unique hart identifier (``hart ID'') that the
{\RISCV} Privileged Architecture assigns to the hart.
Two different interrupt domains may employ entirely different index
numbers for the same set of harts.

\begin{commentary}
For efficiency, implementations should prefer small integers for hart
index numbers.
If an interrupt domain has $H$ harts, the maximum hart index number in
that domain should preferably be no more than around $\mbox{2}H$ or the
system's largest unique {\RISCV} hart ID, whichever is larger.
\end{commentary}

@@

For machine-level interrupt domains configured in MSI delivery mode
(register \z{domaincfg} has field DM =~1), registers \z{mmsiaddrcfg}
and \z{mmsiaddrcfgh} may optionally provide parameters used to
determine the addresses to write outgoing MSIs.

For interrupt domains that are not at machine level or do not support
MSI delivery mode (\z{domaincfg}.DM is read-only zero), these registers
are not implemented, and the eight bytes at their locations are simply
read-only zeros like other reserved bytes.

@@

All fields other than L are {\WARL}.
An implementation is free to choose what values are supported.
Typically, some bits are writable while others are read-only constants.
In the extreme, the values of all fields may be entirely
constant, fixed by the implementation.

If bit~L in \z{mmsiaddrcfgh} is set to one, \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} are \emph{locked}, and writes to the registers are
ignored, making the registers effectively read-only.
When L~=~1, the other fields in \z{mmsiaddrcfg} and \z{mmsiaddrcfgh}
may optionally read as zeros.
In that case, if these other fields were given nonzero values
when L was first set, their values are retained internally by the
PLIC but become no longer visible by reading \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh}.

At system reset, L~is initialized to either zero or one, whichever is
deemed appropriate for the specific PLIC implementation.
If reset initializes L to one, either the other fields are
hardwired by the PLIC to constants, or the PLIC has a different means,
outside of this standard, for determining the addresses of outgoing MSI
writes.
In the latter case, the other fields in \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} may read as zeros, so registers \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} have only read-only values zero and \z{0x80000000}
respectively.
Any time \z{mmsiaddrcfg} or \z{mmsiaddrcfgh} has a different value
(not zero or \z{0x80000000} respectively), the addresses for
outgoing MSI writes directed to machine level must be derivable
from the visible values of these registers, as specified in
Section~\ref{sec:AdvPLIC-MSIAddrs}.

@@

It is expected that most Advanced PLICs will implement a relatively
simple hierarchy of interrupt domains, with the physical addresses
of target IMSICs typically arranged in a linear progression for each
domain.
Registers \z{mmsiaddrcfg} and \z{mmsiaddrcfgh} (along with
\z{smsiaddrcfg} and \z{smsiaddrcfgh} from the next subsection) allow
sufficiently trusted machine-level software, early after system reset,
to configure the pattern of physical addresses for target IMSICs and
then lock this configuration against subsequent tampering.

@@

If an implementation supports additional forms of reset for the
PLIC besides system reset, it is implementation-defined (or possibly
platform-defined) how these other resets may affect \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh}.
However, it must not be possible for insufficiently privileged software
to use a localized reset to unlock these registers by changing bit~L
back to zero.
For this reason, it is likely that only a complete system reset affects
these registers, and any other resets do not.

@@

For machine-level interrupt domains, registers \z{smsiaddrcfg} and
\z{smsiaddrcfgh} may optionally provide parameters used by child
supervisor-level domains to determine the addresses to write outgoing
MSIs.

Registers \z{smsiaddrcfg} and \z{smsiaddrcfgh} are implemented only if:
(a)~the interrupt domain is at machine level;
(b)~the domain's harts implement supervisor mode; and
(c)~the domain has one or more child supervisor-level domains that
support MSI delivery mode (\z{domaincfg}.DM is not read-only zero in at
least one of the supervisor-level child domains).
If the registers are not implemented, the eight bytes at their
locations are simply read-only zeros like other reserved bytes.

The addresses that a supervisor-level interrupt domain uses for MSIs
can be configured only at its parent machine-level domain.

When implemented, \z{smsiaddrcfg} has this format:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 31:0 & Low Base PPN (\WARL) \\
\end{displayLinesTable}
and \z{smsiaddrcfgh} has this format:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bit 31     & L \\
bits 28:24 & HHXS (\WARL) \\
bits 22:20 & LHXS (\WARL) \\
bits 18:16 & HHXW (\WARL) \\
bits 15:12 & LHXW (\WARL) \\
bits 11:0  & High Base PPN (\WARL) \\
\end{displayLinesTable}
All other bits of \z{smsiaddrcfgh} are reserved and read as zeros.

Fields High Base PPN from \z{smsiaddrcfgh} and Low Base PPN from
\z{smsiaddrcfg} concatenate to form a \mbox{44-bit} Base PPN
(Physical Page Number).
The use of this value and fields HHXS (High Hart Index Shift),
LHXS (Low Hart Index Shift), HHXW (High Hart Index Width),
and LHXW (Low Hart Index Width) for
determining target addresses for MSIs is described later, in
Section~\ref{sec:AdvPLIC-MSIAddrs}.

All fields other than L are {\WARL}.
An implementation is free to choose what values are supported, just as
for \z{mmsiaddrcfg} and \z{mmsiaddrcfgh}.

If bit~L in \z{smsiaddrcfgh} is set to one, \z{smsiaddrcfg} and
\z{smsiaddrcfgh} are \emph{locked}, read-only.
When L~=~1, the other fields in \z{smsiaddrcfg} and \z{smsiaddrcfgh}
may optionally read as zeros.
In that case, if these other fields were given nonzero values
when L was first set, their values are retained internally by the
PLIC but become no longer visible by reading \z{smsiaddrcfg} and
\z{smsiaddrcfgh}.

At system reset, L~is initialized to either zero or one, whichever is
deemed appropriate for the specific PLIC implementation.
If reset initializes L to one, either the other fields are
hardwired by the PLIC to constants, or the PLIC has a different means,
outside of this standard, for determining the addresses of outgoing MSI
writes for supervisor-level child domains of this domain.
In the latter case, the other fields in \z{smsiaddrcfg} and
\z{smsiaddrcfgh} may read as zeros, so registers \z{smsiaddrcfg} and
\z{smsiaddrcfgh} have only read-only values zero and \z{0x80000000}
respectively.
Any time \z{smsiaddrcfg} or \z{smsiaddrcfgh} has a different value
(not zero or \z{0x80000000} respectively), the addresses for outgoing
MSI writes from supervisor-level child domains must be derivable
from the visible values of these registers, as specified in
Section~\ref{sec:AdvPLIC-MSIAddrs}.

If an implementation supports additional forms of reset for the
PLIC besides system reset, it is implementation-defined (or possibly
platform-defined) how these other resets may affect \z{smsiaddrcfg} and
\z{smsiaddrcfgh}.
However, it must not be possible for insufficiently privileged software
to use a localized reset to unlock these registers by changing bit~L
back to zero.
For this reason, it is likely that only a complete system reset affects
these registers, and any other resets do not.

@@

@@

To forward interrupts by MSIs, a PLIC must know the MSI target address
for each hart.
For any given system, these addresses are fixed and should be hardwired
into the PLIC if possible.
However, some PLIC implementations may require that software supply the
MSI target addresses.
In that case, registers \z{mmsiaddrcfg},
\z{mmsiaddrcfgh}, \z{smsiaddrcfg}, and \z{smsiaddrcfgh}
(Sections \ref{sec:AdvPLIC-reg-mmsiaddrcfg}
and~\ref{sec:AdvPLIC-reg-smsiaddrcfg}) may be used to configure the
MSI addresses for a machine-level domain and its supervisor-level child
domains.
Alternatively MSI addresses may be configured by some custom means
outside this standard.
If MSI target addresses must be configured by software, this should
be done only from a suitably privileged execution mode, typically just
once, early after system reset.

@@

For a supervisor-level domain, if MSI target addresses are determined
by the parent domain's \z{smsiaddrcfg} and \z{smsiaddrcfgh} registers,
then to construct the address for an outgoing MSI for interrupt
source~$i$, the Hart Index from register \z{target[$i$]} must first be
converted into the index that the machine-level parent domain uses for
the same hart.
(Typically, these numbers are the same, but they may not be.)
The address for the MSI is then computed using this machine-level hart
index, the parent domain's \z{smsiaddrcfg} and \z{smsiaddrcfgh}, and
the Guest Index value from \z{target[$i$]}:
\begin{displayLinesTable}
$g =
  (\mbox{machine-level hart index\z{>>}LHXW})
    \mbox{ \z{\&} }(\mbox{2}^{\rm HHXW}-\mbox{1})$\\
$h = \mbox{machine-level hart index \z{\&} }(\mbox{2}^{\rm LHXW}-\mbox{1})$\\
$\mbox{MSI address} =
  \bigl(\,
    \mbox{Base PPN \z{|} }(g\mbox{\z{<<}}(\mbox{HHXS}+\mbox{12}))
      \mbox{ \z{|} }(h\mbox{\z{<<}LHXS})\mbox{ \z{|} Guest Index}
  \,\bigr)\mbox{\z{<<}12}$
\end{displayLinesTable}

>>>>

Within a given interrupt domain, each of the domain's harts has a
unique \emph{index number} in the range 0 to ${\mbox{2}^{14}-\mbox{1}}$
(=~16,383).
The index number a domain associates with a hart may or may not have
any relationship to the unique hart identifier (``hart ID'') that the
{\RISCV} Privileged Architecture assigns to the hart.
Two different interrupt domains may employ entirely different index
numbers for the same set of harts.
However, if any of a PLIC's interrupt domains can forward interrupts by
MSI, then all machine-level domains of the PLIC share a common mapping
of index numbers to harts.

\begin{commentary}
For efficiency, implementations should prefer small integers for hart
index numbers.
\end{commentary}

@@

For machine-level interrupt domains, registers \z{mmsiaddrcfg}
and \z{mmsiaddrcfgh} may optionally provide parameters used to
determine the addresses to write outgoing MSIs.

If no interrupt domain of the PLIC supports MSI delivery
mode (\z{domaincfg}.DM is read-only zero for all domains),
these two registers are not implemented for any domain.
Otherwise, they are implemented for the root domain, and may
or may not be implemented for other machine-level domains.
For domains not at machine level, they are never implemented.
When a domain does not implement \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh}, the eight bytes at their locations
are simply read-only zeros like other reserved bytes.

Registers \z{mmsiaddrcfg} and \z{mmsiaddrcfgh} are
potentially writable only for the root domain.
For all other machine-level domains that implement them,
they are read-only.

@@

When \z{mmsiaddrcfg} and \z{mmsiaddrcfgh} are writable
(root domain only), all fields other than L are {\WARL}.
An implementation is free to choose what values are supported.
Typically, some bits are writable while others are read-only constants.
In the extreme, the values of all fields may be entirely
constant, fixed by the implementation.

If bit~L in \z{mmsiaddrcfgh} is set to one, \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} are \emph{locked}, and writes to the registers are
ignored, making the registers effectively read-only.
When L~=~1, the other fields in \z{mmsiaddrcfg} and \z{mmsiaddrcfgh}
may optionally all read as zeros.
In that case, if these other fields were given nonzero values
when L was first set in the root domain,
their values are retained internally by the
PLIC but become no longer visible by reading \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh}.

Setting \z{mmsiaddrcfgh}.L to one also locks registers
\z{smsiaddrcfg} and \z{smsiaddrcfgh} described in the next
subsection, if those registers are implemented as well.

For the root domain, L~is initialized at
system reset to either zero or one, whichever is
deemed appropriate for the specific PLIC implementation.
If reset initializes L to one, either the other fields are
hardwired by the PLIC to constants, or the PLIC has a different means,
outside of this standard, for determining the addresses of outgoing MSI
writes.
In the latter case, the other fields in \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} may all read as zeros, so registers \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} have only read-only values zero and \z{0x80000000}
respectively.
Any time \z{mmsiaddrcfg} or \z{mmsiaddrcfgh} has a different value
(not zero or \z{0x80000000} respectively), the addresses for
outgoing MSI writes directed to machine level must be derivable
from the visible values of these registers, as specified in
Section~\ref{sec:AdvPLIC-MSIAddrs}.

For machine-level domains that are not the root domain, if these
registers are implemented, bit~L is always one, and the other
fields either are read-only copies of \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} from the root domain, or are all zeros.

@@

It is expected that most systems will arrange the
physical addresses of target IMSICs in a simple
linear correspondence with hart index numbers.
Registers \z{mmsiaddrcfg} and \z{mmsiaddrcfgh} (along with
\z{smsiaddrcfg} and \z{smsiaddrcfgh} from the next subsection) allow
sufficiently trusted machine-level software, early after system reset,
to configure the pattern of physical addresses for target IMSICs and
then lock this configuration against subsequent tampering.

@@

If a PLIC supports additional forms of reset
besides system reset, it is implementation-defined (or possibly
platform-defined) how these other resets may affect \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} (as well as \z{smsiaddrcfg} and \z{smsiaddrcfgh})
in the root domain.
However, it must not be possible for insufficiently privileged software
to use a localized reset to unlock these registers by changing bit~L
back to zero.
For this reason, it is likely that only a complete system reset affects
these registers, and any other resets do not.

@@

For machine-level interrupt domains, registers \z{smsiaddrcfg} and
\z{smsiaddrcfgh} may optionally provide parameters used by
supervisor-level domains to determine the addresses to write outgoing
MSIs.

Registers \z{smsiaddrcfg} and \z{smsiaddrcfgh} are implemented
by a domain if the domain implements \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} and the PLIC has at least one supervisor-level
interrupt domain.
If the registers are not implemented, the eight bytes at their
locations are simply read-only zeros like other reserved bytes.

Like \z{mmsiaddrcfg} and \z{mmsiaddrcfgh}, registers
\z{smsiaddrcfg} and \z{smsiaddrcfgh} are potentially
writable only for the root domain.
For all other machine-level domains that implement them,
they are read-only.

When implemented, \z{smsiaddrcfg} has this format:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 31:0 & Low Base PPN (\WARL) \\
\end{displayLinesTable}
and \z{smsiaddrcfgh} has this format:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 22:20 & LHXS (\WARL) \\
bits 11:0  & High Base PPN (\WARL) \\
\end{displayLinesTable}
All other bits of \z{smsiaddrcfgh} are reserved and read as zeros.

Fields High Base PPN from \z{smsiaddrcfgh} and Low Base PPN from
\z{smsiaddrcfg} concatenate to form a \mbox{44-bit} Base PPN
(Physical Page Number).
The use of this value and field LHXS (Low Hart Index Shift) for
determining target addresses for MSIs is described later, in
Section~\ref{sec:AdvPLIC-MSIAddrs}.

When \z{smsiaddrcfg} and \z{smsiaddrcfgh} are writable
(root domain only), all fields are {\WARL}.
An implementation is free to choose what values are supported, just as
for \z{mmsiaddrcfg} and \z{mmsiaddrcfgh}.

If register \z{mmsiaddrcfgh} of the domain has bit~L set to one,
then \z{smsiaddrcfg} and \z{smsiaddrcfgh} are \emph{locked}
as read-only alongside \z{mmsiaddrcfg} and \z{mmsiaddrcfgh}.
When \z{mmsiaddrcfgh}.L~=~1, if the readable values of \z{mmsiaddrcfg}
and \z{mmsiaddrcfgh} are zero and \z{0x80000000} respectively---because
their other fields are hidden---then \z{smsiaddrcfg} and
\z{smsiaddrcfgh} are hidden also and read as zeros.

For the root domain only, if \z{mmsiaddrcfgh}.L =~1 and the
MSI-address-configuration fields are hidden (so \z{mmsiaddrcfgh}
reads as \z{0x80000000} and registers \z{mmsiaddrcfg},
\z{smsiaddrcfg}, and \z{smsiaddrcfgh} all read as zeros),
then whatever values \z{smsiaddrcfg} and \z{smsiaddrcfgh} had when
\z{mmsiaddrcfgh}.L was first set are retained internally by the PLIC,
though those values are no longer visible by reading the registers.
Alternatively, if system reset initializes \z{mmsiaddrcfgh}.L =~1 in
the root domain, and if all MSI-address-configuration fields never
appear as anything other than zeros, then the PLIC implementation
has some other, possibly nonstandard, means for determining
the addresses of outgoing MSIs, as discussed in the previous
subsection, \ref{sec:AdvPLIC-reg-mmsiaddrcfg}.

Any time \z{mmsiaddrcfg} and \z{mmsiaddrcfgh} are not read-only
zero and \z{0x80000000} respectively, the addresses for outgoing
MSI writes directed to supervisor level must be derivable from the
visible values of registers \z{mmsiaddrcfgh}, \z{smsiaddrcfg}, and
\z{smsiaddrcfgh}, as specified in Section~\ref{sec:AdvPLIC-MSIAddrs}.

For machine-level domains that are not the root domain,
if \z{smsiaddrcfg} and \z{smsiaddrcfgh} are implemented
and are not read-only zeros, then they are read-only
copies of the same registers from the root domain.

@@

Extempore MSIs are not affected by the IE bit of the domain's
\z{domaincfg} register.
An extempore MSI is sent even if \z{domaincfg}.IE~=~0.

@@

To forward interrupts by MSIs, a PLIC must know the MSI target address
for each hart.
For any given system, these addresses are fixed and should be hardwired
into the PLIC if possible.
However, some PLIC implementations may require that software supply the
MSI target addresses.
In that case, the root domain's registers \z{mmsiaddrcfg},
\z{mmsiaddrcfgh}, \z{smsiaddrcfg}, and \z{smsiaddrcfgh}
(Sections \ref{sec:AdvPLIC-reg-mmsiaddrcfg}
and~\ref{sec:AdvPLIC-reg-smsiaddrcfg}) may be used to configure the
MSI addresses for all interrupt domains.
Alternatively MSI addresses may be configured by some custom means
outside this standard.
If MSI target addresses must be configured by software, this should
be done only from a suitably privileged execution mode, typically just
once, early after system reset.

@@

For a supervisor-level domain, if MSI target addresses are determined
by the root domain's configuration registers (\z{smsiaddrcfg} and others),
then to construct the address for an outgoing MSI for interrupt
source~$i$, the Hart Index from register \z{target[$i$]} must first be
converted into the index number that machine-level domains use for
the same hart.
(These numbers are often the same, but they may not be.)
The address for the MSI is then computed using this machine-level hart
index together with the Base PPN and LHXS values
from \z{smsiaddrcfg} and \z{smsiaddrcfgh},
the other fields (HHXW, LHXW, and HHXS) from \z{mmsiaddrcfgh},
and the Guest Index from \z{target[$i$]}, as follows:
\begin{displayLinesTable}
$g =
  (\mbox{machine-level hart index\z{>>}LHXW})
    \mbox{ \z{\&} }(\mbox{2}^{\rm HHXW}-\mbox{1})$\\
$h = \mbox{machine-level hart index \z{\&} }(\mbox{2}^{\rm LHXW}-\mbox{1})$\\
$\mbox{MSI address} =
  \bigl(\,
    \mbox{Base PPN \z{|} }(g\mbox{\z{<<}}(\mbox{HHXS}+\mbox{12}))
      \mbox{ \z{|} }(h\mbox{\z{<<}LHXS})\mbox{ \z{|} Guest Index}
  \,\bigr)\mbox{\z{<<}12}$
\end{displayLinesTable}
